Multi-level modelling with Openflexo/FML - A contribution to the \mpc

### 2021-05-25 3-Introduction.tex

TODO

\cite{multiProcessChallenge2019}

\emph{Submissions responding to the challenge should describe
a multi-level model conforming to the case description,
including justifications for non-trivial design decisions. In
order to foster comparability between solutions, respondents
are asked to make sure that concepts of the case description
are explicitly represented by one or more model elements.}\\

\emph{criteria:
\begin{enumerate}
    \item Does the response address the established domain as described in the challenge description and demonstrate the use of modeling features?
    \item Does it evaluate/discuss its proposed modeling solution against the criteria described in the challenge description?
    \item Does it discuss the merits and limitations of the employed modeling technique?
    Authors are invited to suggest further requirements to demonstrate particular benefits of the approaches they adopt.
\end{enumerate}
}

structure : plus ou moins imposée par le challenge (structure actuelle = sections au moins attendues)

>>>>

TODO
MDE has traditionally adopted a two-level,top-down meta-modelling approach, where meta-models reside in a certain meta-level, and models are created one meta-level below by using types from the metamodel (cite MOF and EMF).

This strict approach shows its limitations when modeling complex domains which require more than one level of specialization, e.g., to adapt the model to application sub domains. Indeed, the two-level approach fails to acknowledge the existence of two different forms of classification (ontological vs linguistic) and the duality class-object, leading to the introduction of accidental complexity when dealing with the aforementioned complex domains.

In order to tackle this problem multi-level modeling have been introduced. Contrary to the strict approach, multi-level modeling advocates the use of a flexible number of levels as well as a flexiblzation of the instantiation concept.

This approach is gaining interest lately within the modeling community as shown by the MULTI level-modeling workshop and the daghtul seminar and number of tools. Taking advantage of this vibrant state of affairs, and in order to foster discussion and enable comparison between competing approaches, a multi-level modeling challenge has been created. This paper is a response to the latest multi-level modeling challenge which consist of modeling processes. We do so by using model federation and the OpenFlexo framework.

Model federation is a modeling approach in which...

Our solution, which is fully implemented and executable, meets all the requirements.

The rest of the paper is organized as follows.

% \cite{multiProcessChallenge2019}

% \emph{Submissions responding to the challenge should describe
% a multi-level model conforming to the case description,
% including justifications for non-trivial design decisions. In
% order to foster comparability between solutions, respondents
% are asked to make sure that concepts of the case description
% are explicitly represented by one or more model elements.}\\

% \emph{criteria:
% \begin{enumerate}
%     \item Does the response address the established domain as described in the challenge description and demonstrate the use of modeling features?
%     \item Does it evaluate/discuss its proposed modeling solution against the criteria described in the challenge description?
%     \item Does it discuss the merits and limitations of the employed modeling technique?
%     Authors are invited to suggest further requirements to demonstrate particular benefits of the approaches they adopt.
% \end{enumerate}
% }

% structure : plus ou moins imposée par le challenge (structure actuelle = sections au moins attendues)

### 2021-05-25 6-ModelPresentation.tex

@@

Figure \ref{fig:AcmeFullArchitecture} shows the full architecture capturing Acme software development process, with an instance of enacted Software Engineering Process, and highlights all conceptual levels.

@@

>>>>

\todo{Gérer également la fin de P9, non traité pour le moment: artifact they produced must be associated with a validation task}

@@

@@

\todo{Augmenter la taille des fontes de la figure \ref{fig:AcmeFullArchitecture}}

Figure \ref{fig:AcmeFullArchitecture} shows the full architecture capturing Acme software development process, with an instance of enacted \textit{Software Engineering Process}, and highlights all conceptual levels. Acme metamodel enriches base metamodel by offering specializing concepts: \textit{SETaskType} extending \textit{TaskType}, a more specialized concept \textit{CodingTaskType} extending \textit{SETaskType}, \textit{SEArtifactType} extending \textit{ArtifactType}, a more specialized concept \textit{CodeArtifactType} extending \textit{SEArtifactType}. Acme metamodel also defines task type \textit{Coding} as a specialized concept of \textit{Task}, and provides concept \textit{Developer} extending \textit{ActorType}. This metamodel is completed with \textit{ProgrammingLanguage} concept, defined as an enumeration (\textit{Java}, \textit{C}, \textit{COBOL}). All instances required to capture challenge use case are still defined in final Acme model itself instance of Acme metamodel. Instances are represented with rounded boxes and linguistic instantiation are represented with dashed connectors.

Figure \ref{fig:AcmeSoftwareDevelopmentProcess} shows a screen-capture from the tooling developed in the context of the challenge, and presents the \textit{Software Development Process} for Acme company.

\begin{figure}
 \centering
    \includegraphics[width=1.0 \columnwidth]{Figures/SoftwareEngineeringProcessCroped.pdf}
     \caption{Acme software development process}
    \label{fig:AcmeSoftwareDevelopmentProcess}
\end{figure}

\todo{Supprimer ombres et dégradés de la figure \ref{fig:AcmeSoftwareDevelopmentProcess}, et faire en sorte qu'elle soit plus haute et moins large}

\textit{"Requirements analysis"} is defined as an instance of \textit{SETaskType}, \textit{Analyst} as an instance of \textit{ActorType} and \textit{Requirement specifications} as an instance of \textit{SEArtifactType}. \textit{"Requirements analysis"} instance defines property values \textit{producedArtifactTypes=\{Requirement specifications\}} and \textit{allowedActorTypes=\{Analyst\}} (\textbf{S1}). In the same way, \textit{"Test case design"} is defined as an instance of \textit{SETaskType}, marked as critical with property value \textit{isCritical=\{true\}}, and bound to \textit{Analyst} with \textit{allowedActorTypes} relation. \textit{"Test case design"} produces \textit{"Test cases"} defined as an instance of \textit{SETaskType}. The latter is used in "Test design review" (\textit{producedArtifactTypes} relation) (\textbf{S1} and \textbf{S13}, satisfied with \textbf{P9}).

The capture of \textbf{S3} requirement is performed while defining \textit{Coding} as an instance of concept \textit{CodingTaskType} extending \textit{SETaskType} and defining a relation \textit{languages} to \textit{ProgrammingLanguage} with 1..* cardinality. \textit{newTask(String)} behaviour overrides generic implementation while instantiating a \textit{Coding} concept, specializing \textit{SEArtifact}, as shown in following excerpt of FML code:

\begin{lstlisting}[breaklines=true, language=java, basicstyle=\ttfamily\scriptsize, mathescape=true]
concept CodingTaskType extends SETaskType 
{
  ProgrammingLanguage[1..*] languages;
  ...
  public Process newTask(String name) {    
    return new Coding(name,this); 
  }
}    
\end{lstlisting}

\textit{Code} concept extends \textit{SEArtifact}, and defines a relation \textit{language} to \textit{ProgrammingLanguage} with single cardinality (\textbf{S4}). \textbf{S5} is more ambigous as task type \textit{Coding} defines one or more programming languages but produces code which is expressed in one language only. Semantics is unclear relatively to instantiation policy for artifacts. We assume that a Code artifact must be produced for each programming language defined in a \textit{CodingTaskType}. If \textit{COBOL} is referenced in languages defined for a given Coding instance, then we must instantiate a \textit{Code} instance with \textit{COBOL} language (\textbf{S5} and \textbf{S6}).

\begin{lstlisting}[breaklines=true, language=java, basicstyle=\ttfamily\scriptsize, mathescape=true]
concept CodingTaskType extends SETaskType 
{
  ...
  public Artifact[] instantiateArtifacts() {
    Artifact[] returned = {}; 
    for (langage : languages) {
      returned.add(new Code(langage); 
    }
    return returned;
  }
  ...
}    
\end{lstlisting}
